---
title: "datefixR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{datefixR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(datefixR)
```

This vignette describes the functionality of `datefixR` in more detail than the
[README](https://github.com/ropensci/datefixR). `DatefixR` is a lightweight
package consisting of a two user-accessible functions, `fix_date_char()` and
`fix_date_df()`, which converts dates written in different formats into R's
built-in `Date` class. The former is designed to modify character vectors whilst
the latter is intended for modifying columns of a data frame (or tibble).

Practically, this package is most useful when handling date data which has been
supplied via text boxes (instead of a date-specific input with a consistent date
format). However, this package may also be useful to validate the format of date
data (see [date and month imputation](#date-and-month-imputation)).

## Usage

### Date standardization

Firstly, we will demonstrate date standardization without imputation. We
consider a data frame with two columns of dates in differing formats
with no missing data. 

```{R, echo = TRUE}
bad.dates <- data.frame(
  id = seq(5),
  some.dates = c(
    "02/05/92",
    "01-04-2020",
    "1996/05/01",
    "2020-05-01",
    "02-04-96"
  ),
  some.more.dates = c(
    "01 03 2015",
    "02/05/00",
    "01/05/1990",
    "03-Dec-2012",
    "02 April 2020"
  )
)
knitr::kable(bad.dates)
```

`fix_date_df()` requires two arguments, `df`, a data frame (or tibble) object
and `col.names`, a character vector containing the names of columns to be
standardized. By default, the first column of the data frame is assumed to
contain row IDs. These IDs are used if a warning or error is raised to assist
with locating the source of the error. The ID column can also be manually
provided via the `id` argument. 

The output from this function is a data frame object with the selected date
columns now belonging to the `Date` class. 

```{R}
fixed.dates <- fix_date_df(
  bad.dates,
  c("some.dates", "some.more.dates")
)
knitr::kable(fixed.dates)
```

`datefixR` can handle many different formats including -, /, or white space
separation,  year-first or day-first, and month supplied as a number, an
abbreviation or full length name.

`fix_date_char()` is similar to `fix_date_df()` but only applies to a single
character object. 

```{R}
fix_date_char("01 02 2014")
```

Functions in `datefixR` assume day-first instead of month-first when day, month,
and year are all given (unless year is given first). However this behavior can
be modified by passing `format = "mdy"` to function calls.  


```{R}
fix_date_char("01 02 2014", format = "mdy")
```


### Date and month imputation

By default, `datefixR` imputes missing months as July, and missing days of the
month as the first day.  As such, "1992" converts to 

```{R}
fix_date_char("1992")
```

The argument for defaulting to July is 1-2 July is halfway through the year
(on a non leap year). Therefore assuming the year supplied is indeed correct,
you are only a maximum of 6 months off from the true date.  However, this
behavior can be changed by supplying the `day.impute` and  `month.impute`
arguments with an integer corresponding to the desired day and month.
For example, `day.impute = 1` and `month.impute = 1`  results in the first day
of January being imputed instead. 

```{R}
fix_date_char("1992", day.impute = 1, month.impute = 1)
```

The imputation mechanism can also be modified to impute `NA` if a month or day
is missing by setting `day.impute` or `month.impute` to `NA`. This will
also result in a warning being raised.

```{R}
fix_date_char("1992", month.impute = NA)
```

Finally, imputation can be prevented by setting `day.impute` or `month.impute`
to `NULL`. This will result in an error being raised if the day or month
are missing respectively. 

```{R, eval = FALSE}
fix_date_char("1992", month.impute = NULL)
# ERROR
```

`day.impute` and `month.impute` can also be passed to `fix_date_df()` for
similar functionality. 

```{R}
example.df <- data.frame(
  id = seq(1, 3),
  some.dates = c("2014", "April 1990", "Mar 19")
)
fix_date_df(example.df, "some.dates", day.impute = 1, month.impute = 1)
```

## Citation

If you use this package in your research, please consider citing `datefixR`.
An up-to-date citation can be obtained by running

```{R}
citation("datefixR")
```
